Using raw accelerometer data to predict high-impact mechanical loading

### 2022-10-31 manuscript/manuscript.Rmd
Revise manuscript

Bone responds to mechanical loading by altering its properties, such as density, geometry, and strength [@Turner2003; @Fonseca2014; @Santos2017]. Physical activity is a way of eliciting mechanical stimuli, and it is well established to improve bone health. However, there is no consensus in which characteristics of the physical activity-related mechanical stimulus are better associated with bone health benefits. Nevertheless, there is evidence that high-impact exercise, although performed scarcely throughout the day, may play an important role in inducing positive bone adaptations [@Vainionpaa2006; @Stiles2017].

But, even when knowing precisely which the best loading regimen is to improve bone health, the question of how to measure and monitor this load remains. Bone mechanical loading is usually measured in laboratory settings, using instruments such as force plates (FP) [@Medved2000]. Measurements in these instruments, that are confined to a laboratory space, are not applicable in controlling the loads that a person is routinely subjected to in free-living contexts. To surpass this issue, in the last few years, methods using wearable devices such as accelerometers have been proposed to measure mechanical loading, more precisely,  ground reaction force (GRF) [@Neugebauer2012; @Neugebauer2014; @Fortune2014; @Neugebauer2018; @Veras2020] and loading rate (LR) [@Fortune2014; @Veras2020]. These devices are light, small, can be used throughout days, and have a relatively affordable price. These characteristics make the accelerometers well-suited instruments to measure mechanical loading on a daily basis.

However, the validity of the accelerometer-predicted mechanical loading values depends on some factors, as the sample characteristics [@Welk2005]. In fact, due to this, several calibration studies have been done in several populations, such as children [@Neugebauer2012], adults [@Neugebauer2014; @Fortune2014], military personnel [@Neugebauer2018], and subjects with obesity [@Veras2020]. Another important factor that influences the validity of the predictions is the type of activities performed in the calibration study [@Bassett2012]. Until now, studies that calibrate accelerometers to predict mechanical loading variables have usually employed activities such as walking [@Fortune2014; @Neugebauer2018; @Veras2020], or walking and running [@Neugebauer2012; @Neugebauer2014]. Nevertheless, activities eliciting higher impact loads have not yet been tested.

Therefore, the objective of this study was to develop accelerometry-based mechanical loading prediction equations for high-impact activities in subjects from normal weight to severe obesity.

@@

A total of `r max(sample_size$pGRF_N)` adults were enrolled in the study (`r unname(sample_size_sex["M"])` males; `r sample_descriptives$age_mean`±`r sample_descriptives$age_sd` years; `r sample_descriptives$height_mean`±`r sample_descriptives$height_sd` cm; `r sample_descriptives$body_mass_mean`±`r sample_descriptives$body_mass_sd` kg; `r sample_descriptives$BMI_mean`±`r sample_descriptives$BMI_sd` kg·m^-2^, X̄±SD). The protocol was approved by the local Ethics Committee (CES 192-14), and all participants were informed about the experiments purpose and protocol before giving written informed consent. No participant reported any musculoskeletal limitations. The measurement of height (stadiometer model 213; Seca, Hamburg, Germany) and body mass (digital scale model 899; Seca, Hamburg, Germany) was accomplished according to standard practice.

In this study, jumps were chosen as the high-impact activities to be performed. After familiarization, participants completed a series of trials involving jumps of different types and heights. First, drop jumps were executed from steps varying from 5cm to 40cm high, with 5cm increments. Then, box jumps were performed from the floor to 5cm, 15cm, and 30cm high boxes. Finally, continuous jumps were carried out at two different heights: 5cm and 15cm. Drop and box jumps were executed in two sets of four jumps at each height, with 30s interval, and the continuous jumps in one set of 20s at each height. The landing phase of the drop jumps, the take-off phase of the box jumps, and both phases of the continuous jumps were recorded in FC (1000Hz; AMTI Corporation, Watertown, USA). Participants were instructed to jump with no restrictions to arms movement, land two-footed, and then remain stationary on the force plate for five seconds (except in the continuous jumps).

Throughout the protocol, three activity monitors (GT9X Link; 100Hz; ±16g range; ActiGraph, Pensacola, USA) were worn by each participant, at the following placements: i) at their right hip (along the anterior axillary line, at the level of the iliac crest); ii) at their lower back (at the midpoint between the two posterior superior iliac spines); and iii) at their right ankle (immediately superior to the lateral malleolus). Monitors worn at hip and lower back were positioned on the same elastic belt with clips, and the ankle-worn monitor was fixed by an elastic belt and adhesive tape. Activity monitors were always positioned to ensure the alignment of the accelerometer vertical axis and the standing body craniocaudal axis.

The activity monitors used include a primary and a secondary triaxial accelerometers. As the manufacturer proprietary filter is not applied on the secondary accelerometer raw data, only data from this accelerometer were used in this study, to improve data processing replicability. Data from FP and accelerometers were operated through the manufacturer supplied software (for the FP: Netforce, Version 3.5.1; AMTI Corporation, Watertown, USA; for the accelerometers: ActiLife version 6.13.3; Actigraph, Pensacola, USA) and exported as raw data from the *x*, *y* and *z* vectors. GRF was expressed in N and acceleration (ACC) in gravitational ACC units (1*g* = 9.807 m·s^-2^).

The MATLAB software (Version 2019a, Mathworks, Natick, USA) was used to process data from both FP and accelerometers. Data were processed according to the following procedures: first, the GRF signal was resampled to match the activity monitors sampling frequency (100Hz). Then, both GRF and ACC signals were filtered using a Butterworth fourth-order low-pass filter, with 20 Hz cut-off frequency, to attenuate the noise magnitude. This cut-off frequency was selected based on visual inspection of the signals frequency spectrum after a Fast Fourier Transform (FFT). Afterwards, the GRF and ACC resultant vector was calculated ($r_i = \sqrt{x_i^2 + y_i^2 + z_i^2}$). After that, both signals were adjusted using the time set by the systems clock and then synchronized by the maximum cross-correlation coefficient, with posterior manual adjustments if necessary. Next, peak ACC (pACC) was determined by the following criteria: i) a minimum height of three standard deviations above the mean ACC value recorded during a trial; and ii) a separation of at least 0.2 seconds for continuous jumps and 4 seconds for the other jumps. Peak GRF (pGRF) was then defined as the highest value between 0.2 or 4 seconds before and after each pACC, according to the jump type. The rates of change were computed through a centered derivative from the beginning of the foot contact to the curve peak (Equation 1). The peak LR (pLR) and peak acceleration rate (pAR) were defined as the maximum value in the array (Equation 2).

@@

Finally, pGRF, pACC, pLR and pAR means of the resultant and its vertical component for each participant at each jump type and height for each accelerometer placement were extracted and used in all remaining analyses.

Statistical analyses were performed through the R statistical software (version `r paste(R.Version()$major, R.Version()$minor, sep = ".")`, R Foundation for Statistical Computing, Vienna, Austria). All code used to generate the results can be found in an open platform (https://github.com/verasls/jumping_GRF_ACC)(ref). Statistical significance was set as $\alpha$ = 0.05.

Linear mixed models were used to develop pGRF and pLR prediction equations for each vector (resultant and vertical) and accelerometer placement (ankle, lower back, and  hip). Body mass was tested as a predictor in all models, along with the acceleration magnitude (pACC) in the pGRF models and its rate (pAR) in the pLR models. These predictors were entered as fixed effects and kept in the model, as all have shown to be significant (*p* < 0.05). Subject and the interaction between jump type and height were tested as random effects, and both factors have shown to improve the models. Final models were chosen according to the -2 log-likelihood statistic [@Field2012]. The conditional coefficient of determination (R^2^), which estimates the variance explained by both fixed and random factors [@Nakagawa2013], was computed.

Model validation was performed using the leave-one-out cross-validation [@Staudenmayer2012]. This method consists in separating data from one of the subjects into a testing dataset, and keeping the remainder in the training dataset. Then, a new model is developed using the training dataset, with the same parameters that were defined using the whole sample. This new model is then applied to predict the outcome for the subject in the testing dataset. The whole process is repeated once for each participant (`r max(sample_size$pGRF_N)` times). Data from the testing dataset were used in all subsequent analyses.

The quality of the models was then evaluated both numerically and graphically. The numerical assessment was made by computing the following accuracy indices: the mean absolute error (MAE), mean absolute percent error (MAPE) and root mean square error (RMSE). Despite the absence of a standard criterion for establishing an acceptable error in mechanical loading prediction, we regarded a MAPE <8\% to be an accurate prediction based on past research [@Fortune2014; @Neugebauer2014; @Ancillao2018, @Veras2020]. To visualize the agreement between the values predicted by the models and measured by the FP, Bland-Altman plots were drawn. This type of plot is made by plotting the mean of the actual and predicted values on the x axis and their difference on the y axis. Then, the bias was calculated as the mean of these differences and plotted as a continuous horizontal line, and the limits of agreement were defined by bias ± 1.96 standard deviation and plotted as horizontal dashed lines [@Bland1986]. To further test if there were any consistent under- or over-estimation of the values by the models, one-sample *t*-tests were done to check whether the bias were different than zero. Finally, linear regressions were employed to examine if the difference between actual and predicted values were affected by their magnitude, verifying, therefore, whether the prediction was constant throughout the magnitude range that was tested [@Giavarina2015].

Lastly, our pGRF and pLR prediction models were compared with a previously published equation by our group [@Veras2020]. As, for the best of our knowledge, there is no mechanical loading prediction equation developed for jumping activities, we chose to compare it with previous models from our group that were validated in a similar sample. The models used in the comparison were the ones predicting pGRF and pLR from both the resultant and vertical vectors of accelerometers worn at lower back and hip. The MAE, MAPE and RMSE were calculated using the values predicted using Veras et al. [@Veras2020] equation.

@@

In our protocol, three distinct jump types were performed, in several different heights. This variety of jump trials produced a varying response in both the FP- and accelerometer-measured variables. These responses in pGRF, pLR, pACC, and pAR, regarding their resultant vector, can be observed in Figure 1 (see Supplemental Figure S1 for the vertical vector). Panels A and B show the FP-measured variables normalized by the subjects body mass, the pGRF and pLR, respectively. It can be seen that in the drop jump trials, there is a tendency to increase the loading magnitude the higher the jump, and in the box jumps this tendency is lower. As for the continuous jumps, both assayed heights elicit reasonably constant loadings. The same pattern can be observed in the accelerometer-measured variables, pACC and pAR, in Panels C and D, respectively. Also, regarding the influence of the accelerometer placement, in most of the trials accelerometers worn at hip and lower back produced approximately similar values, with a tendency to higher dispersion in data derived from lower back accelerometers.

```{r accuracy_table, results = "show"}
source(here("tables", "tab1.R"))
tab1
```

Accelerometer-based mechanical loading prediction equations were developed based on linear mixed models. The regression coefficients, R^2^ and accuracy indices of the models derived from accelerometers worn at ankle, lower back, and hip, and from the resultant and vertical vectors are presented in Table 1. The R^2^ values for the pGRF models ranged from `r min(GRF_accuracy$R2_cond)` to `r max(GRF_accuracy$R2_cond)`, showing that the models can explain at least `r round(min(GRF_accuracy$R2_cond), 1) * 100`\% of the pGRF variance. The observed MAPE values were between `r myround(min(GRF_accuracy$MAPE) * 100, 1)`\% and `r myround(max(GRF_accuracy$MAPE) * 100, 1)`\%, which falls outside of the proposed threshold of 8\%. Nevertheless, these MAPE values are similar to the results usually found in accelerometer-based energy expenditure prediction [@Diniz-Sousa2020], which have widespread use in research. Despite similar accuracy results achieved by all equations, data from the resultant vector of accelerometers at the hip have shown the lower accuracy error, represented by the smallest MAE (`r myround(hip_GRF_accuracy$MAE, 1)` N), MAPE (`r myround(hip_GRF_accuracy$MAPE * 100, 1)`\%), and RMSE (`r myround(hip_GRF_accuracy$RMSE, 1)` N). Compared to the pGRF models, the pLR equations yielded poorer results, which is expected based on past research [@Fortune2014; @Veras2020]. Their R^2^ ranged from `r min(LR_accuracy$R2_cond)` to `r max(LR_accuracy$R2_cond)` and MAPE from `r myround(min(LR_accuracy$MAPE) * 100, 1)`\% to `r myround(max(LR_accuracy$MAPE) * 100, 1)`\%. However, similar to the pGRF models, the best pLR prediction equation in terms of prediction accuracy was the one from the resultant vector of hip-worn accelerometers, with a MAE of `r myround(hip_LR_accuracy$MAE, 1)` N, MAPE of `r myround(hip_LR_accuracy$MAPE * 100, 1)`\%, and RMSE of `r myround(hip_LR_accuracy$RMSE, 1)` N.

@@

To visualize the agreement between the pGRF and pLR values measured by the FP and predicted by the models, a series of Bland-Altman plots were built for both the resultant and vertical vectors and for all accelerometer placements. Figure 2 shows the Bland-Altman plots for the pGRF (Panel A) and pLR (Panel B) resultant vector of hip-worn accelerometers. Both plots have shown good agreement levels, as at least `r ba_agreement`\% of the data points are within the limits of agreement and data tended to cluster around zero. Bland-Altman plots for the other accelerometer placements and vectors have shown the same patterns described previously, and can be found in the Supplemental Figure S2. Also, as could be observed by the one-sample *t*-tests results, none of the equations presented a bias different than zero (*p* > 0.05). Furthermore, all of the equations have shown a proportional bias, i.e., the magnitude of the values influences the prediction error. However, as the linear regressions presented a very small R^2^ (maximum value of `r myround(max(ba_regression_ankle_R2), 2)`) this effect is neglectable.

```{r accuracy_table2, results = "show"}
source(here("tables", "tab2.R"))
tab2
```

\begin{figure}[ht]
	\captionsetup{singlelinecheck = false, format = hang, justification = raggedright, font = footnotesize, labelsep = space}
	\centering
	\begin{measuredfigure}
		\includegraphics[width=\textwidth]{fig3.png}
    \caption{A comparison of the relationship between peak acceleration and peak ground reaction force (Panel A) and peak acceleration rate and peak loading rate (Panel B) in three ways: i) as measured by the force plates (black round points); ii) predicted by the equations developed in the current study (blue triangle points); and iii) predicted by the equations developed for walking by Veras et al (red square points). Abbreviations: pACC, peak acceleration; pAR, peak acceleration rate; pGRF, peak ground reaction force; pLR, peak loading rate.}
	\end{measuredfigure}
\end{figure}

Although the accuracy of our equations showed to be beyond the established 8\% threshold, when applying a model not developed for jumping activities, we can see that the accuracy tremendously decreases, reaching values of MAPE around `r floor(mean(as.numeric(substr(tab2_df$MAPE, 1, nchar(tab2_df$MAPE) - 2))[1:4]) / 50) * 50`\% for pGRF prediction and around `r floor(mean(as.numeric(substr(tab2_df$MAPE, 1, nchar(tab2_df$MAPE) - 2))[5:8]) / 10) * 10`\% for pLR prediction as can be observed in Table 2. Furthermore, applying to our accelerometer-recorded jumping data a model that is not specific for this activity yields unreasonable results, such as the negative values for pGRF that can be observed in Figure 3, Panel A.

Finally, it is important to mention that an R package was developed [@Veras2021] to make it easier to apply all the necessary data processing steps to use the prediction equations. The package website (https://lveras.com/impactr) has all the installation instructions and documentation necessary to apply this method on your own data.

@@

This study aimed to develop accelerometer-based mechanical loading prediction equations for jumps of diverse types and heights, testing different accelerometer placements and using both the resultant and vertical vectors. The results have shown that jumping pGRF prediction models based on accelerometry data have a lower accuracy compared to models developed for other activities, such as walking and running. Despite that, these models are the correct choice for this kind of activity eliciting hight levels of impact, as applying a model developed for lower impact activities in a sample similar to the current study yielded far worse results in terms of prediction accuracy. Also, as expected, the pLR prediction models performed poorly compared to the pGRF equations.

Bone is known to increase its tissue mass and enhance its micro-architecture arrangement in response to mechanical stimulation [@Turner2003; @Fonseca2014], particularly at higher magnitudes [@Vainionpaa2006; @Stiles2017]. This type of high-intensity impact activities have shown to be more associated with bone health improvements, even if they are usually performed in low quantities throughout the day. In fact, Vainionpää et al [@Vainionpaa2006] demonstrated that performing a relatively low number (<100) of pACC above 4.9*g* daily is positively associated with increases in bone mineral density. Usually, this acceleration magnitude can be achieved mostly by ballistic movements such as jumps [@Vainionpaa2006; @Rowlands2012]. Nevertheless, accelerometry-based mechanical loading regression models have mostly been validated with a set of activities, such as walking and running, that elicit relatively low acceleration magnitudes, usually ranging from approximately 1.2 to 6*g* [@Neugebauer2014; @Veras2020].

The use of a prediction model based on accelerometer data should ideally be done in conditions similar to the study that validated the model, specially in terms of sample characteristics and type and intensity of the activities performed [@Welk2005]. Therefore, if we want to accurately predict the mechanical loading caused by high-impact activities, we should choose a prediction model that has been validated using these activities. Even though the cross-validation accuracy of such a model was poorer compared with other models developed for activities such as walking and running, as demonstrated by our results, the use of these models is crucial to achieve an acceptable prediction accuracy. It was demonstrated by testing a model previously developed by our group [@Veras2020] in a similar sample, but on walking, on the accelerometer-recorded jumping data from the current study. The accuracy indices yielded by applying these models have greatly worsened and these equations also produced some abnormal values, mainly in the pGRF model, in which it predicted negative values of force. These negative pGRF values were mainly as a result of a predictor, the second-degree polynomial of the acceleration, possessing a negative coefficient. While this coefficient captured the plateauing of pGRF as pACC magnitude augments during walking, effectively improving the model [@Veras2020] for walking, this same pattern is not present in the pGRF vs. pACC relationship during jumping, and, therefore, only impairs the accuracy of these equations when applyed to this kind of activity.

>>>>

Bone responds to mechanical loading by altering its properties, such as mass, microarchitecture, and geometry, which ultimately dictate its strength and resistance to fracture [@Turner2003; @Fonseca2014; @Santos2017]. In fact, physical activity is a paramount element to maintain and improve bone health throughout the life course [@Kohrt2004]. However, the mechanical loading osteogenic effect is highly dependent of the physical activity characteristics [@Burr2002]. Several animal studies [@Turner1998; @Turner2003; @Turner2005] helped to form our understanding of the principles that drive the bone tissue adaptative response. In general terms, high-impact dynamic loading, applied at high frequencies and including rest periods seems to favor increases in bone strength [@Turner2003]. This seems to corroborate with the available evidence from humans. For instance, findings from different types of sports have shown that athletes practicing modalities involving high- and/or odd-impact (e.g., volleyball and football) have a greater bone strength than compared to repetitive and low-impact sports (e.g., cycling and swimming) [@Nikander2010]. Nevertheless, the precise way in which the manipulation of mechanical loading variables influences the bone response in humans is not fully understand, particularly due to a difficulty in accurately assessing bone mechanical loading and its components.

Some questionnaires have been developed to capture bone-stimulating characteristics of specific activities [@Groothausen1997; @Weeks2008], but, although the scores generated by such subjective methods seem to be related to bone status, this association appears to be limited and inconsistent across studies [@Farr2011; @Kim2016; @Kim2018]. A more robust approach to assess mechanical loading is the use of equipment such as force plates, the gold standard for ground reaction forces measurement [@Medved2000]. However, the use of force plates is less applicable in some clinical and research settings, as it is an immobile and expensive device, limited to laboratory conditions. In the last few yers, objective methods based on wearable devices have been proposed as a solution to overcome these limitations, using acceleration data to estimate mechanical loading components, typically assessed only through force plates, namely peak ground reaction forces (pGRF) [@Neugebauer2012; @Neugebauer2014; @Fortune2014; @Neugebauer2018; @Veras2020; @Veras2022] and peak loading rate (pLR) [@Fortune2014; @Veras2020; @Veras2022]. Although accelerometry-based equations have shown a high prediction accuracy [@Veras2020; @Veras2022], this versatile mechanical loading assessment approach is not widely used in either clinical practice or research yet. One of the issues that may contribute to this fact is the lack of validation studies covering a broader spectrum of activities, namely those known to have a greater osteogenic effect, such as high-impact activities.

Until now, studies that calibrate accelerometers to predict mechanical loading variables have usually employed only simple activities such as walking [@Fortune2014; @Neugebauer2018; @Veras2020] or walking and running [@Neugebauer2012; @Neugebauer2014; @Veras2022], while activities eliciting higher impact loads have not yet been tested. The fact that these accelerometry-based prediction equations are only valid for a limited set of activities hinders their ability to accurately quantify the amount of loading a person is submitted to daily, as the physical activities performed by a subject are often composed of other types of activity than just walking and running. Therefore, the validity of these methods should be expanded to other activities that are commonly used to induce positive bone adaptations. Jumping exercises, for instance, are frequently used as a strategy to improve bone health in post-menopausal women [@Du2021] and elder men [@Allison2015]. The direct measurement of the mechanical loading involved in these activities would greatly improve our ability to precisely control the loading regimen elicited by these physical exercises and to establish a more reliable relationship between mechanical stimulation and bone anabolic response and, consequently, to establish a clearer relationship between exercise dose and the extent of the bone adaptive response [@Hansford2022]. Therefore, the objective of this study was to develop accelerometry-based mechanical loading prediction equations for high-impact activities in adult subjects with a broad range of body masses, from normal weight to severe obesity.

@@

A total of `r max(sample_size$pGRF_N)` adults were enrolled in the study (`r unname(sample_size_sex["M"])` males; `r sample_descriptives$age_mean`±`r sample_descriptives$age_sd` years; `r sample_descriptives$height_mean`±`r sample_descriptives$height_sd` cm; `r sample_descriptives$body_mass_mean`±`r sample_descriptives$body_mass_sd` kg; `r sample_descriptives$BMI_mean`±`r sample_descriptives$BMI_sd` kg·m^-2^, X̄±SD). The protocol was approved by the local Ethics Committee (CES 192-14), and all participants were informed about the experiments purpose and protocol before giving written informed consent. No participant reported any musculoskeletal limitations. The measurement of height (stadiometer model 213; Seca, Hamburg, Germany) and body mass (digital scale model 899; Seca, Hamburg, Germany) was accomplished according to standard procedures.

In this study, jumps were chosen as the high-impact activity to be performed. After familiarization with the protocol, participants completed a series of trials involving jumps of different types and heights. First, drop jumps were executed from steps varying from 5cm to 40cm high, with 5cm increments. Then, box jumps were performed starting from the floor up to 5cm, 15cm, and 30cm high boxes. Finally, continuous jumps were carried out at two different heights: 5cm and 15cm. Drop and box jumps were executed in two sets of four jumps at each height, with 30s interval, and the continuous jumps in one set of 20s at each height. The landing phase of the drop jumps, the take-off phase of the box jumps, and both phases of the continuous jumps were recorded in a force plate (1000Hz; AMTI Corporation, Watertown, USA). Participants were instructed to jump with no restrictions regarding arms movements, to land two-footed and to remain stationary on the force plate for five seconds after landing (except for the continuous jumps).

Throughout the protocol, three activity monitors (GT9X Link; 100Hz; ±16g range; ActiGraph, Pensacola, USA) were worn by each participant, at the following placements: i) at their right hip (along the anterior axillary line, at the level of the iliac crest); ii) at their lower back (at the midpoint between the two posterior superior iliac spines); and iii) at their right ankle (immediately superior to the lateral malleolus). Monitors worn at the hip and lower back were positioned on the same elastic belt with clips, and the ankle-worn monitor was fixed by an elastic belt and adhesive tape. Activity monitors were always positioned to ensure the alignment of the accelerometer vertical axis and the standing body longitudinal axis. The hip and lower back placements were chosen due to their proximity to the body center of mass, and the ankle placement was chosen as its location is closer to the ground contact point.

The activity monitors used include a primary and a secondary triaxial accelerometers. As the manufacturer proprietary filter is not applied on the secondary accelerometer raw data, only data from this accelerometer were used in this study, to improve data processing replicability. Data from force plates and accelerometers were operated through the manufacturer supplied software (for the force plates: Netforce, Version 3.5.1; AMTI Corporation, Watertown, USA; for the accelerometers: ActiLife version 6.13.3; Actigraph, Pensacola, USA) and exported as raw data from the x, y and z vectors. GRF was expressed in Newton (N) and acceleration in gravitational acceleration units (1*g* = 9.807 m·s^-2^).

The MATLAB software (Version 2019a, Mathworks, Natick, USA) was used to process data from both FP and accelerometers. Data were processed according to the following procedures: first, the ground reaction force signal was resampled to match the activity monitors sampling frequency (100Hz). Then, both ground reaction force and acceleration signals were filtered using a Butterworth fourth-order low-pass filter, with 20 Hz cut-off frequency, to attenuate the noise magnitude. This cut-off frequency was selected based on visual inspection of the signals frequency spectrum after a Fast Fourier Transform (FFT). Afterwards, the ground reaction force and acceleration resultant vector was calculated ($r_i = \sqrt{x_i^2 + y_i^2 + z_i^2}$). After that, both signals were adjusted using the time set by the systems clock and then synchronized by the maximum cross-correlation coefficient, with posterior manual adjustments if necessary. Next, peak acceleration (pACC) was determined by the following criteria: i) a minimum height of three standard deviations above the mean ACC value recorded during a trial; and ii) a separation of at least 0.2 seconds for continuous jumps and 4 seconds for the other jumps. Then, pGRF was then defined as the highest value between 0.2 or 4 seconds before and after each pACC, according to the jump type. The rates of change were computed through a centered derivative from the beginning of the foot contact to the curve peak (Equation 1). The pLR and peak acceleration rate (pAR) were defined as the maximum value in the array (Equation 2).

@@

Finally, pGRF, pACC, pLR and pAR means of the resultant vector and its vertical component for each participant at each jump type and height for each accelerometer placement were extracted and used in all remaining analyses.

Statistical analyses were performed through the R statistical software (version `r paste(R.Version()$major, R.Version()$minor, sep = ".")`, R Foundation for Statistical Computing, Vienna, Austria). All code used to generate the results can be found in an open platform (https://bit.ly/3szNSJS). Statistical significance was set as $\alpha$ = 0.05.

Linear mixed models were used to develop pGRF and pLR prediction equations for each vector (resultant and vertical) and accelerometer placement (ankle, lower back, and hip). Body mass was tested as a predictor in all models, along with the acceleration magnitude (pACC) in the pGRF models and its rate (pAR) in the pLR models. These predictors were entered as fixed effects and kept in the model, as all were shown to be significant (*p* < 0.05). Subject and the interaction between jump type and jump height were tested as random effects, and both factors were shown to improve the models. Final models were chosen according to the -2 log-likelihood statistic [@Field2012]. The conditional coefficient of determination (R^2^), which estimates the variance explained by both fixed and random factors [@Nakagawa2013], was computed.

Model validation was performed using the leave-one-out cross-validation [@Staudenmayer2012]. This method consists in separating data from one of the subjects into a testing dataset and keeping the remainder in the training dataset. Then, a new model is developed using the training dataset, with the same parameters that were defined using the entire sample. This new model is then applied to predict the outcome for the subject in the testing dataset. The whole process is repeated once for each participant (`r max(sample_size$pGRF_N)` times). Data from the testing dataset were used in all subsequent analyses.

The quality of the models was then evaluated both numerically and graphically. The numerical assessment was made by computing the following accuracy indices: the mean absolute error (MAE), mean absolute percent error (MAPE) and root mean square error (RMSE). To visually analyze the agreement between the values predicted by the models and measured by the force plates, Bland-Altman plots were drawn. These plots are constructed by plotting the mean of the actual and predicted values on the x axis and their difference on the y axis. Then, the bias was calculated as the mean of these differences and plotted as a continuous horizontal line, and the limits of agreement were defined by bias ± 1.96 standard deviation and plotted as horizontal dashed lines [@Bland1986]. To further test if there were any systematic under- or over-estimation of the values by the models, one-sample *t*-tests were done to check whether the bias were significantly different from zero. Finally, linear regressions were employed to examine if the difference between actual and predicted values were affected by their magnitude, verifying, therefore, whether the prediction was constant throughout the entire magnitude range [@Giavarina2015].

@@

In our protocol, three distinct jump types were performed, each one using several different heights. This variety of jump trials produced a considerable heterogeneity in the response of both the force plates- and accelerometer-measured variables. The results obtained for pGRF, pLR, pACC, and pAR, regarding their resultant vector, can be observed in Figure 1 (see Supplemental Figure S1 for the vertical vector). Panels A and B show the force plates-measured variables normalized by the subject’s body mass, the pGRF and pLR, respectively. It can be seen that, in the drop jump trials, there is a tendency for the loading magnitude to increase as the jump height increases. This tendency is also observed for the box jumps, but with a lower magnitude. As for the continuous jumps, both assayed heights elicited reasonably constant loadings. Overall, a similar pattern can be observed in the accelerometer-measured variables, pACC and pAR, in panels C and D, respectively. Nevertheless, this pattern is less clear, particularly for the continuous jumps. Regarding the influence of the accelerometer placement on the variables collected, in most of the trial’s, accelerometers worn at the hip and lower back produced approximately similar values, with a tendency for higher dispersion in data derived from the lower back accelerometers.

```{r accuracy_table, results = "show"}
source(here("tables", "tab1.R"))
tab1
```
Accelerometer-based mechanical loading prediction equations were developed based on linear mixed models. The regression coefficients, R^2^ and accuracy indices of the models derived from accelerometers worn at the ankle, lower back and hip, and from the resultant and vertical vectors are presented in Table 1. The R^2^ values for the pGRF models ranged from `r min(GRF_accuracy$R2_cond)` to `r max(GRF_accuracy$R2_cond)`, showing that all the models were able to explain at least `r round(min(GRF_accuracy$R2_cond), 1) * 100`\% of the pGRF variance, and the observed MAPE values were between `r myround(min(GRF_accuracy$MAPE) * 100, 1)`\% and `r myround(max(GRF_accuracy$MAPE) * 100, 1)`\%. Despite similar accuracy results achieved by all equations, the model derived from data from the resultant vector of accelerometers worn at the hip were shown to have the lower accuracy error, represented by the smallest MAE (`r myround(hip_GRF_accuracy$MAE, 1)` N), MAPE (`r myround(hip_GRF_accuracy$MAPE * 100, 1)`\%), and RMSE (`r myround(hip_GRF_accuracy$RMSE, 1)` N).

Compared to the pGRF models, the pLR equations yielded poorer results. Although their R^2^ ranged from `r min(LR_accuracy$R2_cond)` to `r max(LR_accuracy$R2_cond)`, error measures were high, with MAPE ranging from `r myround(min(LR_accuracy$MAPE) * 100, 1)`\% to `r myround(max(LR_accuracy$MAPE) * 100, 1)`\%. Nevertheless, similar to the pGRF models developed, the best pLR prediction equation in terms of prediction accuracy was the one from the resultant vector derived from hip-worn accelerometers, even thought error estimates were higher, with a MAE of `r myround(hip_LR_accuracy$MAE, 1)` N, MAPE of `r myround(hip_LR_accuracy$MAPE * 100, 1)`\%, and RMSE of `r myround(hip_LR_accuracy$RMSE, 1)` N.

@@

To visualize the agreement between the pGRF and pLR values measured by the force plates and predicted by the models, a series of Bland-Altman plots were built for both the resultant and vertical vectors and for all accelerometer placements. Figure 2 shows the Bland-Altman plots for the pGRF (panel A) and pLR (panel B) resultant vector of hip-worn accelerometers. Both plots have shown good agreement levels, as at least `r ba_agreement`\% of the data points are within the limits of agreement and the data tended to cluster around zero. Bland-Altman plots for the other accelerometer placements and vectors have also shown the same previously described patterns and can be found in Supplemental Figure S2. Also, as could be observed by the one-sample *t*-tests results, none of the equations presented a bias significantly different than zero (*p* > 0.05). Furthermore, all of the equations have shown a proportional bias, i.e., the magnitude of the values influences the prediction error. However, as the linear regressions used to assess the proportional bias presented a very small R^2^ (maximum value of `r myround(max(ba_regression_ankle_R2), 2)`), this effect is neglectable.

Finally, an R package was also developed [@Veras2021] to simplify the application of all the necessary data processing steps involved in the implementation of the developed prediction equations. The package website (https://verasls.github.io/impactr) has all the installation instructions and documentation necessary to apply this method on the user’s own data.

@@

This study aimed to develop accelerometer-based mechanical loading prediction equations, namely for pGRF and pLR, for jumping exercises of different types and heights, as well as to test different accelerometer placements and data derived from both the resultant and vertical acceleration vectors. The results have shown that pGRF prediction models for jumping activities based on accelerometry data have a MAPE around 13%, while pLR prediction equations resulted in a lower accuracy, with a MAPE around 22%. Also, these models are the first to be developed specifically for high-impact jumping activities in adults.

Bone is known to increase its tissue mass and enhance its micro-architecture arrangement in response to mechanical stimulation [@Turner2003; @Fonseca2014], particularly of higher magnitudes [@Vainionpaa2006; @Stiles2017]. These types of high-intensity impact activities have shown to be more associated with bone health improvements, even if they are usually performed in low volumes throughout the day. In fact, Vainionpää et al [@Vainionpaa2006] demonstrated that performing a relatively low number (<100) of pACC above 4.9g daily is positively associated with increases in bone mineral density. Usually, this acceleration magnitude can be achieved mostly by specific movements such as jumps [@Vainionpaa2006; @Rowlands2012] . Nevertheless, accelerometry-based mechanical loading regression models have mostly been validated for a set of activities, such as walking and running, that elicit relatively low acceleration magnitudes [@Neugebauer2014; @Veras2020; @Veras2022]. The results of this study provide new mechanical loading prediction models for jumping, which are especially important as it is one of the most popular and effective activities prescribed for bone health improvement [@Beck2017].

The development of such models was done with the direct purpose of allowing to assess the loading sustained by the skeletal system during daily physical activity. Nowadays, physical activity monitorization and exercise prescription aiming bone health promotion has been carried out almost blindly or through the use of subjective instruments such as questionaries. Moreover, the absence of a trustworthy method to monitor mechanical loading on a daily basis may preclude the prescription and self-involvement of patients in high-impact activities that may expose them to a higher risk of joint injuries. These issues may be overcome with the recent popularization of wearable sensors, such as accelerometers, and their ability to be used as a tool for mechanical loading prediction. This would allow some important clinical applications, as the models developed in this study, when used in combination with other prediction equations for other types of activities [@Veras2022], can be used to fully analyze the mechanical loading spectrum that people are submitted to daily and relate to bone health parameters. Therefore, a more complete understanding can be achieved on how mechanical loading influences bone health, by, for instance, identifying the loading volume and magnitude range that are more efficient in inducing bone health benefits. Also, the development and distribution of an R package that automates the steps necessary to apply these models enhances their applicability to potential users [@Veras2021]. This can prove to be an important tool, as the application of these models depend on some laborious steps, such as the application of digital filters to reduce noise from the signal, detection of the peaks in this signal and the computation of derivatives to obtain the acceleration rate, necessary for the pRL prediction.
